---
title: "Reproducible Research Assignment 1"
author: "Ken Ammerman"
date: "Wednesday, Sept 3rd, 2014"
output:
  html_document:
    keep_md: yes
---


## Assignment
To create a single R markdown document that contains the assignment answers 
along with the code used to load, transform, process and visualize date 
obtained from a [Activity Monitoring data](https://d396qusza40orc.cloudfront.net/repdata/data/activity.zip) dataset.


##### Load the data from the .csv extracted to a ./data sub-directory in the 
##### current working directory.

```{r global_variable, echo=TRUE}
library(ggplot2)

srcFile <- "data/activity.csv"
activityData <- read.csv(srcFile, header=TRUE, sep = ",")
```

Display the structure of the date loaded
```{r}
str(activityData)
```

##### Sumarize the step counts to a daily total 

```{r}
dailyData <- aggregate(steps ~ date, activityData, sum)
```

##### Create a histogram of the total number of steps taken per day

```{r}
hist(dailyData$steps, xlab="Daily step totals", 
     main="Total Steps Each Day", col="blue")
```

##### Mean Total numner of steps per day
```{r}
mean(dailyData$steps)
```

##### Median total number of steps each day
```{r}
median(dailyData$steps)
```


##### Plot the average daily activity pattern
```{r}
averageData <- aggregate(steps ~ interval, activityData, mean )
plot(averageData,xlab="5 minute interval",
     ylab="Average steps", type='l',col="blue")
```

To find the 5 minute interval with the maximum average daily steps
```{r}
averageData[which.max(averageData[,2]),]
```



#### Input missing values by using the average daily values for an interval
##### Calculate the Total number of missing values
```{r}
sum(is.na(activityData))
```


#### Replace missing values with the mean of the daily steps.
##### This is not intended to be a sophisticated solution, simply replace the
##### NA values with the average interval value
```{r}
avgSteps <- mean(activityData$steps, na.rm = TRUE)
newActivityData <- activityData
newActivityData = replace(newActivityData, is.na(newActivityData), avgSteps)
```

##### structure of newActivityData
```{r}
str(newActivityData)
```


##### Create a histogram using the average steps values to replace the intervals
##### containing missing steps data
```{r}
newDailyData <- aggregate(steps ~ date, activityData, sum)

hist(newDailyData$steps, xlab="Daily step totals", 
     main="Total Steps Each Day", col="green")
```

##### new mean
```{r}
mean(newDailyData$steps)
```

##### and new median
```{r}
median(newDailyData$steps)
```

#### Changes in activity between weekdays and weekends
##### Create a new variable that identified mid-week and weekend days in the
##### newActivityData data set.
```{r}
newActivityData$day <- ifelse ((weekdays(as.Date(newActivityData$date)) 
                            %in% c('Saturday','Sunday')), 'Weekend', 'Weekday')
```

##### new structure of the updated newActivityData
```{r}
str(newActivityData)
```

##### Plot the average number of steps taken per 5 minute interval across all 
##### days grouped by weekday and weekend
```{r}

qplot(interval, steps, 
      data = aggregate( steps~day+interval, newActivityData, mean ),
      facets = .~day, geom="line")
```